---
title: "LIHC data integration"
output:
    html_notebook: default
    html_document: default
---

# Set up

```{r}
library(FactoMineR)
library(tidyverse)
library(ComplexHeatmap)
library(ggpubr)
library(iClusterPlus)
library(tsne)
#substitue the python path below, with the python used to install SUMO
reticulate::use_python(Sys.which("/home/admin/miniconda3/bin/python"), required=TRUE)
library(reticulate)
library(survival)
library(survminer)
library(factoextra)
library(GGally)
library(ggalluvial)
library(PINSPlus)
```

```{r}
# Specify path to directory with downloaded data
data_dir_path = "./data/liver/"
stopifnot(file.exists(file.path(data_dir_path, 'exp_tr')) & file.exists(file.path(data_dir_path, 'methy')) & 
          file.exists(file.path(data_dir_path, 'mirna_tr')))
```

```{r}
# Inrease plot resolution in this notebook
options(repr.plot.res=200)
```

```{r}
# Load data, leave only primary solid tumor samples:
data_exp <- read.table(file.path(data_dir_path, 'exp_tr')) %>% select(ends_with("01"))
data_met <- read.table(file.path(data_dir_path, 'methy')) %>% select(ends_with("01"))
data_mirna <- read.table(file.path(data_dir_path, 'mirna_tr')) %>% select(ends_with("01"))
subtypes <- read_tsv("./data/TCGA_subtypes/LIHC_subtypes.tsv", show_col_types = FALSE)
```

```{r}
# set color palettes we are going to use for subtype visualization
exp_subtypes_col = setNames(RColorBrewer::brewer.pal(name = "Set2", n = 6),
                            unique(na.omit(subtypes$Subtype_mRNA)))
miRNA_subtypes_col = setNames(RColorBrewer::brewer.pal(name = "Set2", n = 6),
                              unique(na.omit(subtypes$Subtype_miRNA)))
DNAmeth_subtypes_col = setNames(RColorBrewer::brewer.pal(name = "Set2", n = 5),
                                unique(na.omit(subtypes$Subtype_DNAmeth)))
iCluster_subtypes_col = setNames(RColorBrewer::brewer.pal(name = "Set2", n = 4),
                                 unique(na.omit(subtypes$Subtype_iCluster)))
paradigm_subtypes_col = setNames(RColorBrewer::brewer.pal(name = "Set2", n = 5),
                                 unique(na.omit(subtypes$Subtype_Paradigm)))
```

```{r}
bound_matrices <- plyr::rbind.fill(data_exp, data_met, data_mirna) # matrix concatenation
dim(bound_matrices)
```

```{r}
subtypes_data <- tibble(sample_codes=colnames(bound_matrices)) %>%  #in order of columns in bound_matrices
    left_join(subtypes, by = "sample_codes")  %>% 
    mutate(exp = sample_codes %in% colnames(data_exp),
           met = sample_codes %in% colnames(data_met),
           mirna = sample_codes %in% colnames(data_mirna))
head(subtypes_data)
```
# iCluster & iCluster+

```{r}
# 367 in total
# With known iCluster subtype — 183
common_samples <- subtypes_data %>% filter(exp & met & mirna & Subtype_iCluster != "iCluster.NA") %>% 
  pull(sample_codes)
exp.common <- data_exp[, common_samples]
met.common <- data_met[, common_samples]
mirna.common <- data_mirna[, common_samples]
```

```{r}
# let's leave only 5000 the most variable genes, as it's recommended in iManual
most_variable <- names(sort(apply(exp.common, 1, var), decreasing = TRUE)[1:5000])
exp.common.small <- exp.common %>% filter(row.names(exp.common) %in% most_variable)
```

**Here, we should tune the parameters.** Let's skip it for the moment.

## try 3 clusters without any parameter tuning

```{r}
# run the iClusterPlus function on common samples data set (this will take a couple of minutes)
r.icluster <- iClusterPlus(t(exp.common.small), t(met.common), t(mirna.common),
  type=rep("gaussian", 3), # Providing the distributions
  K=2, # as in Subtype_iCluster (number of clusters is k+1)
  alpha=c(1,1,1), # as well as other model parameters
  lambda=c(.98,.98,.98))
```

```{r}
# extract the H and W matrices from the run result
# here, we refer to H as z, to keep with iCluster terminology
icluster.z <- r.icluster$meanZ
rownames(icluster.z) <- colnames(exp.common) # fix the row names
icluster.ws <- r.icluster$beta
```

```{r}
# insight into multiple pairs of factors
as_tibble(icluster.z, rownames="sample_codes") %>% 
    left_join(subtypes_data, by = "sample_codes") %>%
    select(sample_codes, V1, V2, Subtype_iCluster) %>%
    na.omit() %>%
    ggpairs(columns=2:4, ggplot2::aes(col=Subtype_iCluster))
```

```{r}
# We run tSNE for visualization purposes and inspect sample separation by existing subtypes
icp.tsne <- tsne(icluster.z)
icp.tsne.tbl <- as_tibble(icp.tsne) %>%
    rename(tSNE_1=V1, tSNE_2=V2) %>%
    mutate(sample_codes = colnames(exp.common))
```

```{r}
p1 <- icp.tsne.tbl %>%
    left_join(subtypes_data, by = "sample_codes") %>%
    ggplot() + geom_point(aes(x=tSNE_1, y=tSNE_2, color=as.factor(Subtype_iCluster), size=1, alpha=0.5)) + 
    labs(title="tSNE for iCluster+ [iCluster subtypes]", color="iCluster subtype") + guides(alpha=FALSE, size=FALSE) 

p1

icp.tsne.tbl %>%
    left_join(subtypes_data, by = "sample_codes") %>%
    mutate(icp_label=r.icluster$clusters) %>%
    ggplot() + geom_point(aes(x=tSNE_1, y=tSNE_2, color=as.factor(icp_label), size=1, alpha=0.5)) + 
     labs(title="tSNE for iCluster+", color="iCluster+ label") + guides(alpha=FALSE, size=FALSE)
```
So, we see that our separation is not the same at all as the existing data for iCluster subtypes. But, we also do not know which data they integrated. Hmm...

## try tuning parameters

```{r}
# set.seed(123)
# for(k in 1:5){
#   cv.fit = tune.iClusterPlus(cpus=2, t(exp.common.small), t(met.common), t(mirna.common),
#   type=rep("gaussian", 3), K=k,  scale.lambda=c(1,1,1))
#   save(cv.fit, file=file.path(data_dir_path, paste("cv.fit.k", k, ".Rdata", sep="")))
# }
```
Fuh, it takes tooooo long...

## try 5 clusters without any parameter tuning

```{r}
r.icluster <- iClusterPlus(t(exp.common.small), t(met.common), t(mirna.common),
  type=rep("gaussian", 3), # Providing the distributions
  K=4, # (number of clusters is k+1)
  alpha=c(1,1,1), # as well as other model parameters
  lambda=c(.98,.98,.98))
```

```{r}
# extract the H and W matrices from the run result
# here, we refer to H as z, to keep with iCluster terminology
icluster.z <- r.icluster$meanZ
rownames(icluster.z) <- colnames(exp.common) # fix the row names
icluster.ws <- r.icluster$beta
```

```{r}
# insight into multiple pairs of factors
as_tibble(icluster.z, rownames="sample_codes") %>% 
    left_join(subtypes_data, by = "sample_codes") %>%
    select(sample_codes, V1, V2, V3, V4, Subtype_mRNA) %>%
    na.omit() %>%
    ggpairs(columns=2:5, ggplot2::aes(col=Subtype_mRNA))
```

```{r}
# We run tSNE for visualization purposes and inspect sample separation by existing subtypes
icp.tsne <- tsne(icluster.z)
icp.tsne.tbl <- as_tibble(icp.tsne) %>%
    rename(tSNE_1=V1, tSNE_2=V2) %>%
    mutate(sample_codes = colnames(exp.common))
```

```{r}
p1 <- icp.tsne.tbl %>%
    left_join(subtypes_data, by = "sample_codes") %>%
    ggplot() + geom_point(aes(x=tSNE_1, y=tSNE_2, color=as.factor(Subtype_mRNA), size=1, alpha=0.5)) + 
    labs(title="tSNE for iCluster+ [mRNA subtypes]", color="mRNA subtype") + guides(alpha=FALSE, size=FALSE) 
p2 <- icp.tsne.tbl %>%
    left_join(subtypes_data, by = "sample_codes") %>%
    ggplot() + geom_point(aes(x=tSNE_1, y=tSNE_2, color=as.factor(Subtype_miRNA), size=1, alpha=0.5)) + 
    labs(title="tSNE for iCluster+ [miRNA subtypes]", color="miRNA subtype") + guides(alpha=FALSE, size=FALSE) 
p3 <- icp.tsne.tbl %>%
    left_join(subtypes_data, by = "sample_codes") %>%
    ggplot() + geom_point(aes(x=tSNE_1, y=tSNE_2, color=as.factor(Subtype_DNAmeth), size=1, alpha=0.5)) + 
    labs(title="tSNE for iCluster+ [DNAmeth subtypes]", color="DNAmeth subtype") + guides(alpha=FALSE, size=FALSE) 


p1
p2
p3

icp.tsne.tbl %>%
    left_join(subtypes_data, by = "sample_codes") %>%
    mutate(icp_label=r.icluster$clusters) %>%
    ggplot() + geom_point(aes(x=tSNE_1, y=tSNE_2, color=as.factor(icp_label), size=1, alpha=0.5)) + 
     labs(title="tSNE for iCluster+", color="iCluster+ label") + guides(alpha=FALSE, size=FALSE)
```
Hmm...


